\bf The usage of bootstrap method in shape-restricted regression

### 2023-09-11 Manuscript/manuscript.tex
Add files via upload

In certain area and cases, apply shape restrictions, such as monotonicity or convexity to estimate a certain function is reasonable and not uncommon in many fields.\cite{guo2019smooth} There are several methods to build shape-restricted regressions to do analysis the regressions, like quadratic programming.\cite{meyer2013simple} However, due to shape restrictions, it is not quite accurate to estimate the uncertainty of the parameters with normal methods like MLE (Maximum likelihood estimation) since parameters will sometimes lies on the boundary. In this paper, we consider to use bootstrap method to replace MLE method to estimate the uncertainty of the parameters, specifically, of certain time points in the regression curve. Although the bootstrap method is a valid method used to construct confidence intervals for regression coefficients, \cite{efron1985bootstrap}and its effectiveness have been widely recognized in linear regression analysis \cite{efron1979bootstrap}, and non-linear regression analysis \cite{davidson1999bootstrap}. However, when dealing with shape-restricted regressions, where the coefficients are forced to adhere to specific constraints such as monotonicity or curvature, the effectiveness of bootstrap is not proved to be guaranteed. In this paper, we will demonstrate the usage of the bootstrap method in shape-restricted regression and proof the effectiveness of bootstrap for constructing pointwise confidence intervals in shape-restricted regression empirically.

The rest of the paper is organized as the follows. Section~\ref{Bootstrap Method for Constructing Confidence Intervals} gives a review of bootstrap method for constructing confidence intervals. Section~\ref{Shape Restricted Example} gives out a shape-restricted regression example and the usage of bootstrap to build the pointwise confidence intervals for certain time points. Section~\ref{Simulation Study} shows a simulation study to assess the performance of the methods.
explores the case where a combination of the first two scenarios occurs. An  
adjusted bootstrap procedure is proposed as a working solution in this case.  
Section~\ref{Conclusion} concludes with a discussion.

@@

Shape-restricted splines regressions are piecewise polynominal, differentiable to a certain degree and have certain restrictions like monotonicity or convexity.\cite{meyer2008inference} Given these conditions, the application of the bootstrap method to estimate pointwise confidence intervals for specific time points within the regression cannot be guaranteed to be suitable. And in the rest of the article we gives out an example of shape-restricted regression and the method to fit confidence interval for certain time points in the function.

\section{Shape Restricted Example}
\label{Shape Restricted Example}

In some situations the regression is forced to be shape-restricted to improve the predictive performance and reduce overfitting if the underlying regression function takes the specific form. Some regressions are naturally constrained to convexity or concativity in a lot of disciplines like economics, biology, and psychological studies, and other ares.\cite{jieying2022heteroscedastic} \cite{guntuboyina2018nonparametric} Some popular examples like cost function and profit function in economics\cite{gallant1984imposing}, precipitation data analysis\cite{molitor2002bayesian}. In this article we gives out an example of the relationship between income and the age of Canadian workers. The data is given from R package SemiPar. For those situations, Regression splines are smooth, flexible, and parsimonious nonparametric function estimators, and it offers an effective approach to enhance the inference of shape-restricted regression. \cite{meyer2008inference} In this dataset, a nondecreasing relationship between the age of workers and the expected logarithm of the income was considered. \cite{wang2021shape}  Splines2 provides implementation of those spline basis functions via R language, and we use that to build a non-decreasing curve by i-spline that force the coefficients to be non-negative with degree of freedom = 6. Then, we selected 10 time points, denoted as $t$, with equal intervals from the dataset, corresponding to one of them is the regression time points \(f'(t)\) that fitted by iSpline. We also performed bootstrap resampling 1000 times to obtain the 95\% quantile pointwise confidence interval for each points. In the figure below, the red line represents the fitted i-spline regression, the yellow line is the smooth regression, and we build ten pointwise confidence intervals for 10 certain time points. The result is shown in Figure ~\ref{fig:semipar}.

@@

Since the underlying distribution is unknown, we need to perform a simulation study to test whether the fitted pointwise confidence interval is accurate for the regression.

We define three functions: \(y = 0.25(x - 0.9)\) , \(y = 0.02(x^3+1.2)\) , and \(y = 0.1(5\Phi((x - 2)/0.3) + 1)\) , where \(\Phi\) is the cumulative distribution function of the standard normal distribution. For those functions, the validity of the bootstrap method to build the confidence intervals for the coefficients and standard error has already been proved. \cite{jieying2022heteroscedastic} And in this article we want to further prove the validity of the bootstrap method to build the pointwise confidence intervals for certain time points in the domain of the function. First of all, we randomly generate some data points \(x\) in the domain of \([0,4]\), and we calculate its corresponding values \(y(x)\), and make simulation data points \(f'(x)\) by add noise to the true value \(f(x)\). The noise is added from a standard normal distribution with \(\mu = 0\) and \(\sigma = \) \(1\). And then we make the paired values \((x, y\text{ with noise})\) be the dataset, while we know the true values of \(f(x)\) as well.

\begin{figure}[H]
  \centering
  \includegraphics[width=0.8\textwidth]{DataPointsGeneration.png}
  \caption{The generated dataset while n=100, true functions and fitted i-splines regressions: the shaded green curve represents the true functions, the pink points are the data pair \((x, f'(x))\) where \(f'(x)\) is generated by \(f(x)\) add noise, and the red curves are the fitted i-splines regression curve with degrees of freedom\(= 6,10,15\), from top to bottom}
  \label{fig:DataAndRegressions}
\end{figure}

We use quadratic i-splines function to enforce the monotonicity of the function and we picked degrees of freedom 6 and 10 to build the regression for the dataset. The generated dataset and regressions is shown in Figure ~\ref{fig:DataAndRegressions}. While according to the law of large number, we also expect that the result will be closer to \(95\%\) while n gets larger, so we also did a simulation for \(n = 1000\) to see the numeric coverage percentiles. In each configuration, we select \(0.5,1,1.5,2,2.5,3,3.5,4\) as the time points from the dataset and perform bootstrap resampling 1000 times to estimate the \(95\%\) quantile pointwise confidence intervals for each of the selected time points. We repeat this process 1000 times to see how many times these confidence intervals can contain the true result. The results are shown in Table ~\ref{Table} and Table ~/ref{2}.

In table ~\ref{Table1}, \(x\) represents the certain time point we selected in the domain of the function, and \(f(x)\) is its true value calculated by the certain known function. CP represents coverage percentage of pointwise confidence interval we built by the bootstrap methods under certain degrees of freedom. We can see that in all those functions and degrees of freedom we have given, the bootstrap method gives pointwise confidence intervals coverage that are closed to the nominal level \(95\%\). And we can also see that while n gets larger, the result of \(n = 1000\), which can be seen in Table ~\ref{Table2} is much better than \(n = 100\) in each same level of degrees of freedom. While degrees of freedom is also an important variable inside the study and we can see that for the different degrees of freedom the result of CP is different either, but since it is not the main scope of this paper, we will not concern a lot about how to choose the best model to enhence the CP result.

>>>>

In certain area and cases, apply shape restrictions, such as monotonicity or convexity to estimate a certain function is reasonable and not uncommon in many fields.\cite{guo2019smooth} There are a lot of important applications that includes shape-restricted regressions in various fields.  Guntuboyina and Sen gives a great summary of the usage of shape-restricted regressions and how those restrictions(like convexity and concativity) arises in different fields.\cite{guntuboyina2018nonparametric} There are several methods to estimate the shape-restricted regressions and build the model for inference like quadratic programming.\cite{meyer2013simple} However, since the parameters of such curves are depends on degrees of freedom, we are more interested in analysis the uncertainty of the curve itself. Due to shape restrictions, when we want to estimate the uncertainty of the curve,it is not quite accurate to estimate it with normal methods like MLE (Maximum likelihood estimation) since parameters will sometimes lies on the boundary. In this paper, we consider to use the bootstrap method to replace MLE method to estimate the uncertainty of the curve. Since the parameters of such curves are depends on degrees of freedom, we are more interested in analysis the uncertainty of the curve itself, specifically, of certain time points inside the regression curve. Although the bootstrap method is a valid method used to construct confidence intervals for regression coefficients, \cite{efron1985bootstrap}and its effectiveness have been widely recognized in linear regression analysis \cite{efron1979bootstrap}, and non-linear regression analysis \cite{davidson1999bootstrap}. However, when dealing with shape-restricted regressions, where the coefficients are forced to adhere to specific constraints such as monotonicity or curvature, the effectiveness of bootstrap is not proved to be guaranteed. In this paper, we will demonstrate the usage of the bootstrap method in shape-restricted regression and proof the effectiveness of bootstrap for constructing pointwise confidence intervals in shape-restricted regression empirically.

The rest of the paper is organized as the follows. Section~\ref{Bootstrap Method for Constructing Confidence Intervals} gives a review of bootstrap method for constructing confidence intervals. Section~\ref{Simulation Study} shows a simulation study to assess the performance of the bootstrap method in shape-restricted cases. And Section~\ref{Real Data Analysis} gives out a shape-restricted regression example based on a real dataset and the use the bootstrap method to build the pointwise confidence intervals for certain time points and access the method. An
adjusted bootstrap procedure is proposed as a working solution in this case.  
Section~\ref{Conclusion} concludes with a discussion.

@@

In some situations the regression is forced to be shape-restricted to improve the predictive performance and reduce overfitting if the underlying regression function takes the specific form. Some regressions are naturally constrained to convexity or concativity in a lot of disciplines like economics, biology, and psychological studies, and other ares.\cite{jieying2022heteroscedastic} \cite{guntuboyina2018nonparametric} Some popular examples like cost function and profit function in economics\cite{gallant1984imposing}, precipitation data analysis\cite{molitor2002bayesian}.

Shape-restricted splines regressions are piecewise polynominal, differentiable to a certain degree and have certain restrictions like monotonicity or convexity.\cite{meyer2008inference} Given these conditions, the application of the bootstrap method to estimate pointwise confidence intervals for specific time points within the regression cannot be guaranteed to be suitable. So in this article, we are going to give out a simulation study to test the accessibility and accuracy of the bootstrap method for shape-restricted regression situations.

Since in the shape-restricted cases the underlying distribution is unknown, we want to perform a simulation study to test whether the fitted pointwise confidence interval is accurate for the shape- restricted regressions.

To do this, We define three functions: \(y = 0.25(x - 0.9)\) , \(y = 0.02(x^3+1.2)\) , and \(y = 0.1(5\Phi((x - 2)/0.3) + 1)\) , where \(\Phi\) is the cumulative distribution function of the standard normal distribution. For those functions, the validity of the bootstrap method to build the confidence intervals for the coefficients and standard error has already been proved. \cite{jieying2022heteroscedastic} And in this article we want to further prove the validity of the bootstrap method to build the pointwise confidence intervals for certain time points in the domain of the function. First of all, we randomly generate some data points \(x\) in the domain of \([0,4]\), and we calculate its corresponding values \(y(x)\), and make simulation data points \(f'(x)\) by add noise to the true value \(f(x)\). The noise is added from a standard normal distribution with \(\mu = 0\) and \(\sigma = \) \(1\). And then we make the paired values \((x, y\text{ with noise})\) be the dataset, while we know the true values of \(f(x)\) as well.

\begin{figure}[H]
  \centering
  \includegraphics[width=0.8\textwidth]{Regression.jpg}
  \caption{The generated dataset while n=100, true functions and fitted i-splines regressions: the shaded green curve represents the true functions, the pink points are the data pair \((x, f'(x))\) where \(f'(x)\) is generated by \(f(x)\) add noise, and the red curves are the fitted i-splines regression curve with degrees of freedom\(= 6, 10\), from top to bottom}
  \label{fig:DataAndRegressions}
\end{figure}

We use quadratic i-splines function to enforce the monotonicity of the function and we picked degrees of freedom 6 and 10 to build the regression for the dataset. The generated dataset and regressions is shown in Figure ~\ref{fig:DataAndRegressions}. While according to the law of large number, we also expect that the result will be closer to \(95\%\) while n gets larger, so we also did a simulation for \(n = 1000\) to see the numeric coverage percentiles. In each configuration, we select \(0.5,1,1.5,2,2.5,3,3.5,4\) as the time points from the dataset and perform bootstrap resampling 1000 times to estimate the \(95\%\) quantile pointwise confidence intervals for each of the selected time points. We repeat this process 1000 times to see how many times these confidence intervals can contain the true result. The results are shown in Table ~\ref{Table1} and Table ~\ref{Table2}.

In table ~\ref{Table1}, \(x\) represents the certain time point we selected in the domain of the function, and \(f(x)\) is its true value calculated by the certain known function. CP represents coverage percentage of pointwise confidence interval we built by the bootstrap methods under certain degrees of freedom. We can see that in all those functions and degrees of freedom we have given, the bootstrap method gives pointwise confidence intervals coverage that are closed to the nominal level \(95\%\). And we can also see that while n gets larger, the result of \(n = 1000\), which can be seen in Table ~\ref{Table2} is much better than \(n = 100\) in each same level of degrees of freedom. We can also see that in some situations, for example, when df = 10, some empirical coverage rate is as low as 0.3 to 0.4, this may due to the degrees of freedom is not suitable for this data regression, which might cause some problems. We can also see that via Figure ~\ref{Figure1} when degrees of freedom equals to 10, the curve is not that smooth compare to the curve of degrees of freedom 6, and in some situations, its shape is not even closed to the true function. So we can see that degree of freedom is a crucial variable and different degrees of freedom will impact the result of CP of the confidence interval. It should be cautious to choose when we are using the splines regression since it will influence the result of the inference a lot. However, since it is not the main scope of this paper, we will not concern a lot about how to choose the best model to enhence the CP result. Overall, we can see that those pointwise confidence intervals made by bootstrap method works fine in suitable degrees of freedom, and in the rest of this article we will use a real world shape-restricted data to give out an data analysis with the bootstrap method.

\section{Real Data Analysis}
\label{Real Data Analysis}

 In this part we gives out an example of the relationship between income and the age of Canadian workers. The data is given from R package SemiPar. For those situations, Regression splines are smooth, flexible, and parsimonious nonparametric function estimators, and it offers an effective approach to enhance the inference of shape-restricted regression. \cite{meyer2008inference} In this dataset, a nondecreasing relationship between the age of workers and the expected logarithm of the income was considered. \cite{wang2021shape}  Splines2 provides implementation of those spline basis functions via R language, and we use that to build a non-decreasing curve by i-spline that force the coefficients to be non-negative with degree of freedom = 6. Then, we selected 10 time points, denoted as $t$, with equal intervals from the dataset, corresponding to one of them is the regression time points \(f'(t)\) that fitted by iSpline. We also performed bootstrap resampling 1000 times to obtain the 95\% quantile pointwise confidence interval for each points. In the figure below, the red line represents the fitted i-spline regression, the yellow line is the smooth regression, and we build ten pointwise confidence intervals for 10 certain time points. The result is shown in Figure ~\ref{fig:semipar}.

@@
